---
title: "Violent_Property_Crime_EDA"
author: "Ian Sargent"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Packages
```{r}
pacman::p_load(tidyverse, sf, MASS, vdc, GGally, units)
```

### Read the Dataset
```{r}
# Read from GitHub Repository
df <- st_read(dsn = "https://raw.githubusercontent.com/iansargent/DC-Metro-Crime/refs/heads/main/Data/Combined/master.geojson")

census <- read.csv("https://raw.githubusercontent.com/iansargent/DC-Metro-Crime/fa2bcfffb8245fd5a3573c419787e96b840aff0a/Data/Census/Census_ACS.csv") |>
  st_as_sf(crs = 4326, wkt = "WKT") |>
  st_transform(3857) |> 
  mutate(
    sq_miles = st_area(WKT) / 2589988.11,
    pop_density = total_population / sq_miles
  ) |> 
  units::drop_units() |> 
  arrange(by = NAME)

metro <- st_read(dsn = "https://raw.githubusercontent.com/iansargent/DC-Metro-Crime/refs/heads/main/Data/Metro/Metros_Ground_Transfer.geoJson")
```

# Do violent and property crime rates vary differently as a function of transit accessibility across census tracts in Washington, DC?

### Aggregate Specific Crime Counts by Census Tract
```{r}

crime_by_tract <- df |>
  st_drop_geometry() |>
  summarize(
    .by = name,
    violent  = sum(crime_type == "violent",  na.rm = TRUE),
    property = sum(crime_type == "property", na.rm = TRUE),
    population = mean(as.numeric(total_population))
  ) |>
  pivot_longer(
    cols = c(violent, property),
    names_to = "crime_type",
    values_to = "count"
  ) |> 
  arrange(by=name)

crime_by_tract <- crime_by_tract |>
  left_join(
    census |> dplyr::select(NAME, WKT),
    by = c("name" = "NAME")
  ) |>
  arrange(by=name) |> 
  rename("geometry" = "WKT") |> 
  st_as_sf()

crime_by_tract
```


### Count of Metro Stations within a 1000 meter buffer
```{r}
# 1. Centroids
census$centroid <- st_centroid(census$WKT)
centroids <- census |> st_set_geometry("centroid") |> st_transform(3857)

# 2. 1000m buffer
centroids$buffer_1000m <- st_buffer(centroids$WKT, 1000)

# 3. Count stations inside buffer
inter_buffer <- st_intersects(centroids$buffer_1000m, metro |> st_transform(3857))
centroids$stations_within_1000m <- lengths(inter_buffer)

# 4. Merge back to crime_by_tract
crime_by_tract <- crime_by_tract |>
  left_join(
    centroids |> st_drop_geometry() |> dplyr::select(NAME, stations_within_1000m),
    by = c("name" = "NAME")
  ) |> 
  st_drop_geometry() |> 
  merge(census, by.x="name", by.y="NAME") |> 
  distinct()
```


### Variable Distributions

#### Total Population 
```{r, warning=FALSE}
crime_by_tract |> 
  ggplot(
    mapping = aes(
      x = population
    )
  ) +
  geom_histogram(bins=30, aes(y=..density..), fill="steelblue", color="white") +
  geom_density(color="darkblue") +
  theme_minimal() +
  labs(
    title = "Census Tract Population Distribution",
    x = "Population",
    y = "Density"
  )
```


#### Population Density
```{r, warning=FALSE}
crime_by_tract |> 
  ggplot(
    mapping = aes(
      x = pop_density
    )
  ) +
  geom_histogram(bins=30, aes(y=..density..), fill="steelblue", color="white") +
  geom_density(color="darkblue") +
  theme_minimal() +
  labs(
    title = "Census Tract Population Distribution",
    x = "Population Density",
    y = "Density"
  )
```

#### Violent Crimes
```{r, warning=FALSE}
crime_by_tract |>
  filter(crime_type == "violent") |> 
  ggplot(
    mapping = aes(
      x = count
    )
  ) +
  geom_histogram(bins=30, aes(y=..density..), fill="mediumseagreen", color="white") +
  geom_density(color="darkgreen") +
  theme_minimal() +
  labs(
    title = "Census Tract Violent Crime Count Distribution",
    x = "Violent Crimes",
    y = "Density"
  )
```


#### Property Crimes
```{r, warning=FALSE}
crime_by_tract |>
  filter(crime_type == "property") |> 
  ggplot(
    mapping = aes(
      x = count
    )
  ) +
  geom_histogram(bins=30, aes(y=..density..), fill="salmon", color="white") +
  geom_density(color="darkred") +
  theme_minimal() +
  labs(
    title = "Census Tract Property Crime Count Distribution",
    x = "Property Crimes",
    y = "Density"
  )
```


#### Stations within 1000m
```{r, warning=FALSE}
crime_by_tract |>
  ggplot(
    mapping = aes(
      x = stations_within_1000m
    )
  ) +
  geom_histogram(bins=17, aes(y=..density..), fill="mediumpurple", color="white") +
  geom_density(color="darkslateblue") +
  theme_minimal() +
  labs(
    title = "Census Tract Stations Distribution",
    x = "Number of Stations within 1000m",
    y = "Density"
  )
```


```{r, warning=FALSE}
crime_by_tract |> 
  dplyr::select(count, stations_within_1000m, total_population, total_housing_units, median_household_income, poverty_rate, unemployment_rate, pct_bachelors_or_higher, pop_density) |> 
ggpairs()
```


```{r}
# Violent crime
nb_violent_offset <- glm.nb(
  count ~ median_household_income + stations_within_1000m + unemployment_rate + pct_non_hispanic_black + offset(log(total_population)) + total_housing_units + pop_density,
  data = filter(crime_by_tract, crime_type == "violent")
)

summary(nb_violent_offset)
```


```{r}
# Property crime
nb_property_offset <- glm.nb(
  count ~ stations_within_1000m + median_household_income + unemployment_rate + pct_non_hispanic_black + offset(log(total_population)) + pop_density,
  data = filter(crime_by_tract, crime_type == "property")
)

summary(nb_property_offset)
```

```{r}


```

