---
title: "New combined data"
author: "David McLone"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(pacman)
```

```{r}
library(janitor)
```

```{r}
library(dplyr)
library(lubridate)
```


### Packages
```{r}
pacman::p_load(tidyverse, sf)
```

```{r}
crime_files <- tibble::tibble(
  year = c(2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025),
  url = c(
    "https://raw.githubusercontent.com/iansargent/DC-Metro-Crime/main/Data/Crime/Crime_Incidents_in_2017.geojson",
    "https://raw.githubusercontent.com/iansargent/DC-Metro-Crime/main/Data/Crime/Crime_Incidents_in_2018.geojson",
    "https://raw.githubusercontent.com/iansargent/DC-Metro-Crime/main/Data/Crime/Crime_Incidents_in_2019.geojson",
    "https://raw.githubusercontent.com/iansargent/DC-Metro-Crime/main/Data/Crime/Crime_Incidents_in_2020.geojson",
    "https://raw.githubusercontent.com/iansargent/DC-Metro-Crime/main/Data/Crime/Crime_Incidents_in_2021.geojson",
    "https://raw.githubusercontent.com/iansargent/DC-Metro-Crime/main/Data/Crime/Crime_Incidents_in_2022.geojson",
    "https://raw.githubusercontent.com/iansargent/DC-Metro-Crime/main/Data/Crime/Crime_Incidents_in_2023.geojson",
    "https://raw.githubusercontent.com/iansargent/DC-Metro-Crime/main/Data/Crime/Crime_Incidents_in_2024.geojson",
    "https://raw.githubusercontent.com/iansargent/DC-Metro-Crime/fa2bcfffb8245fd5a3573c419787e96b840aff0a/Data/Crime/Crime_Incidents_in_2025.geojson"
  )
)
```

```{r}
crime_list <- purrr::map2(crime_files$year, crime_files$url, \(yr, u) {
  x <- st_read(u, quiet = TRUE)
  x$year <- yr
  x
})

crime <- dplyr::bind_rows(crime_list)
```

```{r}
table(crime$year)
```


### Load Datasets
```{r}
census <- read.csv("https://raw.githubusercontent.com/iansargent/DC-Metro-Crime/fa2bcfffb8245fd5a3573c419787e96b840aff0a/Data/Census/Census_ACS.csv") |>
  st_as_sf(crs = 4326, wkt = "WKT") |>
  st_transform(3857)

metro <- st_read(dsn = "https://raw.githubusercontent.com/iansargent/DC-Metro-Crime/refs/heads/main/Data/Metro/Metros_Ground_Transfer.geoJson")
```

## 1. 2025 Crime Incidents

### Classify Crimes as Violent or Property

In accordance with the FBI's Uniform Crime Reporting (UCR) standards, we have classified homicide, assault with a dangerous weapon, sexual abuse, and robbery as violent offenses. Property offenses include theft from automotive, burglary, motor vehicle theft, arson, and general theft. 

```{r}
# A vector of violent crimes
violent <- c("HOMICIDE", "ASSAULT W/DANGEROUS WEAPON", "SEX ABUSE", "ROBBERY")

# Creating the crime_type column (violent vs property)
crime <-
  crime |>
  mutate(crime_type = if_else(OFFENSE %in% violent, "violent", "property"))
```

### Classify Crimes as Automotive Related

```{r}
# A vector of violent crimes
auto <- c("THEFT F/AUTO", "MOTOR VEHICLE THEFT")

# Creating the crime_type column (violent vs property)
crime <-
  crime |>
  mutate(auto_crime = if_else(OFFENSE %in% auto, 1, 0))
```

## 2. Distance to Nearest Stations

```{r}
# Ensuring same map projection for distance calculations
crime <- st_transform(crime, 3857)
metro <- st_transform(metro, 3857)
census <- st_transform(census, 3857)
```


### Adding Nearest Station and Distance (Meters)
```{r}
crime$nearest_station <- st_nearest_feature(crime, metro)
crime$distance_meters <- as.numeric(st_distance(crime, metro[crime$nearest_station, ], by_element = TRUE))

crime_coords <- st_coordinates(crime)
metro_coords <- st_coordinates(metro)

crime$distance_manhattan <-
  abs(crime_coords[, 1] - metro_coords[crime$nearest_station, 1]) +
  abs(crime_coords[, 2] - metro_coords[crime$nearest_station, 2])
```


```{r}
tract_centroids <- census |> st_centroid()
dist_matrix <- st_distance(tract_centroids, metro)
nearest_dist <- apply(dist_matrix, 1, min)
tract_centroids$tract_dist_to_metro <- as.numeric(nearest_dist)
```

```{r}
census <- census |>
  mutate(tract_dist_to_metro = as.numeric(nearest_dist))
```



## 3. Joining Crime and Metro Datasets
```{r}
# Join crime and metro by unique nearest station ID
crime_joined <- crime |>
  rename("nearest_station_id" = "nearest_station") |>
  left_join(
    metro |>
      st_drop_geometry() |>
      dplyr::select(OBJECTID, NAME, Above_Ground, Transfer_Station),
    by = c("nearest_station_id" = "OBJECTID")
  ) |>
  rename("nearest_station_name" = "NAME")
```


## 4. Incorporate Census ACS Controls (By Tract)
```{r}
master_df <-
  st_join(crime_joined, census) |>
  janitor::clean_names()
```


### Write master dataset to files
```{r}
st_write(master_df, "master.geojson", delete_dsn = TRUE)
master_df <- st_read("master.geojson")
```

```{r}
names(master_df)
```

```{r}
# Setting date format to create new seasonality variable
master_df$start_date <- ymd_hms(master_df$report_date)
```

```{r}
# 1) Parse the date-time column (try common formats)
master_df <- master_df %>%
  mutate(
    start_date_parsed = parse_date_time(
      start_date,
      orders = c("Ymd HMS", "Ymd HM", "Ymd", "mdY HMS", "mdY HM", "mdY"),
      tz = "America/New_York"
    ),
    # 2) Create season from month
    season = case_when(
      month(start_date_parsed) %in% c(12, 1, 2) ~ "Winter",
      month(start_date_parsed) %in% c(3, 4, 5) ~ "Spring",
      month(start_date_parsed) %in% c(6, 7, 8) ~ "Summer",
      month(start_date_parsed) %in% c(9, 10, 11) ~ "Fall",
      TRUE ~ NA_character_
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))
  )

table(master_df$season, useNA = "ifany")
```

```{r}
View(master_df)
```
